---
title: "536: Snow in the Cascades"
author: Aren Lampman
format: pdf
echo: false
---

```{r}
#| include: false
library(tidyverse)
library(dplyr)
library(readr)
```

## Introduction

This report analyzes data collected on the daily snow-water equivalent (SWE) measurements from 1992 to 2022 at two SNOTEL locations in the Cascades: Hog Pass and Santiam Pass. This data is acquired from the USDA Natural Resources Conservation Service National Water and Climate Center Report Generator. The data contains one SWE recording per day for each month, for every station for each year from 1992 to 2022. Note, although the data is related to the calendar year, the analysis henceforth uses the term "year" to refer to the water year, which begins October 1st and ends September 30th of the following year. This report will address the following questions of interest:

1.  **Is there any change in the yearly peak amount of snow over the period 1992–2022?**

2.  **In the 2018 water year, how did the daily values of snow water equivalent compare between the two stations?**

\newpage

## Question of Interest 1: Is there any change in the yearly peak amount of snow over the period 1992–2022?

Towards answering this question, we first determined the water year corresponding to each calendar year. The data was then grouped by the station name and water year, where we then calculated the maximum WSE measurement across the entire water year, for each water year. The max SWE per water year for each station was then portrayed in the following line plot:

```{r}
#| include: false
snow_cascade <- read.csv("data/report.csv", skip = 57, na = c('',""))
snow_cascade <- snow_cascade[-1,]
snow_cascade[,4:15] <- lapply(snow_cascade[,4:15], parse_number)
```

```{r}
#| include: false
snow_cascade_long <- snow_cascade |>
  pivot_longer(c(Jan:Dec), names_to = "Month", values_to = "Snow_Water_Equivalent") |>
  mutate(
    Water_Year = ifelse(
      test = Month == "Oct" | Month == "Nov" | Month == "Dec",
      yes = Year,
      no = Year - 1
      )
    ) |>
  filter(!is.na(Snow_Water_Equivalent))
```

```{r}
#| include: false
snow_cascade_max_SWE <- snow_cascade_long |> 
  group_by(Station.Name, Water_Year) |>
  summarise(max_SWE = max(Snow_Water_Equivalent))
```

```{r}
#| warning: false
#| fig-cap: "Line plot of the maximum SWE measurements per station from the 1992 - 2022 water years."
#| fig-width: 6
#| fig-height: 4
snow_cascade_max_SWE |>
  ggplot(mapping = aes(x = Water_Year, y = max_SWE)) +
  geom_line(aes(colour = Station.Name)) +
  labs(
    title = "Max SWE per Water Year per Station",
    x = "Water Year",
    y = "Max SWE",
    color = "Station Name"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  ) +
  scale_x_continuous(
    limits = c(1992, 2022),
    breaks = seq(1992, 2022, by = 2)
  ) +
  scale_y_continuous(
    limits = c(0, 70),
    breaks = seq(0, 70, by = 5)
  )
```

## Conclusion:

First, measurements of SWE can be related to snow amounts in the following way: a low SWE value indicates that the snow depth/density for that day was low. From the line plot, there is a general decrease in yearly peak amount of snow for both stations, as indicated by the relative decrease in max SWE values each year from 1992 to 2022.\

This is especially prevalent for Hogg Pass Station, where, following a short general increase in max SWE from 1992 (SWE measurement of 48.1) to 1998 (SWE measurement of 68.5), the yearly max SWE follows a general decrease. Specifically, from 1998, the max SWE decreased to a max SWE of 54.0 in 2001, followed by a max SWE of 50.3 in 2007, to a max SWE of 35.5 in 2016. Between these years, the max SWE fluctuates between lower and higher values, but these listed years mark the most extreme peaks in max SWE that are not matched by the max SWE for years following these peaks (**every year following one of these listed peaks exhibits a lower max SWE**). As the SWE values for these isolated years is decreasing (68.5 to 54 to 50.3), there is indication of a general decline in max SWE at Hogg Pass Station. As a low SWE measurement indicates a low snow depth, we see that the yearly peak amount of snow is declining from 1992 to 2022 for Hogg Station.\

Santiam Junction Station also exhibits this general decrease in max SWE from 1992 to 2022, albeit with fewer extreme peaks. For instance, the max SWE reading of 37.2 in 1998 marked the greatest max SWE value for years until 2007, when the max SWE was recorded as 38.4. However, this 9-year gap was marked by a general decrease in max SWE values, and the max SWE values recorded after 2007 until 2022 never returned to the high of 38.4. Instead, this gap is again marked by a period of smaller peak max SWE values, peaking at 24.7 in 2010 and 23.8 in 2016. Thus, Santiam Junction Station also demonstrates a general decline in max SWE (and thus amount of snow) from 1992 to 2022.\

As a result, based on this analysis, from 1992 to 2022, **yearly peak amount of snow for both Hogg Pass Station and Santiam Junction Station exhibited a general decline in amount**. Gap years between max peaks exhibit non-monotone behavior in SWE measurements, but the relative spikes in max SWE readings around certain years (described above) are never again reached with the passing of the water years from 1992 to 2022.\

\newpage

## Question of Interest 2: In the 2018 water year, how did the daily values of snow water equivalent compare between the two stations?

This question is addressed by considering the snow-water equivalent(SWE) measurements for each day in the 2018 water year (October 2018 - September 2019). The analysis looks at the calendar dates, but orders the dates by the water year: this is done to make conclusions based on the first and last days of nonzero SWE values. The daily SWE measurements are plotted on the line graph below, and the first and last days of nonzero SWE measurements are summarized in the table below.

```{r}
#| include: false
snow_cascade_long <- snow_cascade_long |>
  mutate(Calendar_Date = lubridate::ymd(paste(Year, Month, Day))) |>
  mutate(Water_Date = lubridate::ymd(paste(Water_Year, Month, Day)))
```

```{r}
#| include: false
snow_cascade_2018 <- snow_cascade_long |>
  filter(Water_Year == 2018) |>
  arrange(Calendar_Date) |>
  select(Station.Name, Snow_Water_Equivalent, Calendar_Date)
```

```{r}
#| warning: false
#| fig-cap: "Line plot of the daily SWE measurements per station for the 2018 water year(October 1st, 2018 - September 30th, 2019)."
#| fig-width: 6
#| fig-height: 4
snow_cascade_2018 |>
  ggplot(mapping = aes(x = Calendar_Date, y = Snow_Water_Equivalent)) +
  geom_line(aes(colour = Station.Name)) +
  labs(
    title = "Daily Snow-Water Equivalent(SWE) vs Date per Station",
    x = "Date",
    y = "SWE",
    color = "Station Name"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  ) +
  scale_x_date(
    date_breaks = "1 months"
    ) +
  scale_y_continuous(
    limits = c(0, 25),
    breaks = seq(0, 25, by = 5)
  )
```

```{r}
#| include: false
snow_cascade_2018_Filtered <- snow_cascade_2018 |>
  filter(Snow_Water_Equivalent > 0) |>
  group_by(Station.Name) |>
  summarise(
    First_Nonzero_SWE = first(Calendar_Date), 
    Last_Nonzero_SWE = last(Calendar_Date)
    )
```

```{r}
#| warning: false
snow_cascade_2018_Filtered |> 
  knitr::kable(
    col.names = c(
      "Station Name", 
      "First Nonzero SWE Date", 
      "Last Nonzero SWE Date" 
    ),
    digits = c(0, 0, 0),
    caption = c("Summary table with the first and last dates corresponding to nonzero Snow-Water Equivalent(SWE) measurements.")
  )
```

## Conclusion:

First considering the line plot, the daily SWE measurements for 2018 indicate that Hogg Pass Station measures a considerably higher daily SWE reading than Santiam Junction Station. This is shown by the plot for Hogg Pass Station residing entirely above or equal to that of Santiam Junction Station across all days. This trend is consistent for every day of measured SWE values. \

The line plot also shows that the maximum SWE measurement in the 2018 water year for Hogg Pass Station was 24.6, occurring on March 29th, 2019, while the maximum measurement for Santiam Junction Station was 15.9, occurring on March 14th, 2019. These dates are relatively close together, and this trend is present for the other dates as well: in general, both stations follow a similar date pattern in rising and falling SWE measurements throughout the 2018 water year. A significant fall mirrored by both stations occurred following the peak SWE measurements for both stations, starting on April 16th, 2019 for Hogg Pass Station and March 14th, 2019 for Santiam Junction Station. Following these dates, both station's daily SWE measurements declined steadily down to 0.\

The table reports on the first and last dates of nonzero SWE measurements for both stations, for the 2018 water year. Uniquely, Hogg Pass Station its first nonzero SWE reading occur on October 6th, 2018, compared to Santiam Junction Station's first nonzero SWE reading on November 23rd, 2018. This is unique because, considering the line plot again, this early start date for Hogg Pass Station is followed by an extended period of 0 SWE measurements until late November. This suggests that October 6th, 2018 is an outlier SWE measurement for Hogg Pass Station. \

Hogg Pass Station also exhibits a longer nonzero SWE reading timeline than that for Santiam Junction Station, with an ending nonzero SWE measurement date of May 26th, 2019, compared to April 17th, 2019 for Santiam Junction Station. These early start and late end dates for nonzero SWE measurements indicate that Hogg Pass Station has a longer snow cover duration than Santiam Junction Station.\

Ultimately, the data from the line plot and table suggests that, for the 2018 water year, Hogg Pass Station had an earlier first nonzero SWE date and later last nonzero SWE date than Santiam Junction Station, indicating a greater snow cover duration for Hogg Pass Station. The daily SWE measurements were considerably higher for Hogg Pass Station than for Santiam Junction Station throughout the 2018 water year. Finally, the peaks and troughs of daily SWE measurements are generally similar for both Hogg Pass Station and Santiam Junction Station, with both station's max SWE measurements occurring in mid-late March of 2019. This peak is followed by a drastic monotone decline in SWE values for both stations until their respective last nonzero SWE measurement dates.  

